Epistemic Semantics in Guarded String Models

### 2020-06-17 tex/draft.tex
sketch section 4

Rather than guarded strings, elements of a guarded string model for
KAT are sets of guarded strings.  In the application, these elements
have the interpretation of propositions (sets of possible worlds)
and/or event types.  An event such as $u$ in the guarded string model
corresponds to the set of guarded strings where the bare event is
flanked by compatible truth assigments, as defined in an Epik program
by the event declaration and the constraint declaration.  These have
the status of event types, in that they can ``happen'' in different
possible worlds. Here happening corresponds to incrementing the world
with some compatible element of the event type.  Where $e$ is a bare
event symbol, $\llbracket e \rrbracket^{\phi,P}$ is the corresponding
set of guarded strings, consisting of the bare event decorated with
compatible truth assignments denoted by $P$ that are admitted by the
world-constraints $\phi$.  See this formally in
Figure~\ref{fig:interp}, and the examples in \ex{events1}.
% u (!q!p):(!qp) + (!qp):(q!p) + (q!p):(qp) 173
% d (!qp):(!q!p) + (q!p):(!qp) + (qp):(q!p)

\begin{example} \label{events1}
\(
\begin{array}[t]{l|c}
  e & \llbracket e \rrbracket^{\phi, P} \\ \hline  
  u & \{ 00u01, 01u10, 10u11\}\\
  d & \{ 01d00, 10d01, 11d10\}\\
  a_1 & \{10 a_1 10 \}\\
  a_0 & \{01 a_0 01\}\\
  b_1 & \{10 b_1 10 \}\\
  b_0 & \{01 b_0 01 \}\\
\end{array}
\)
\end{example}

@@

We interpret each term in as a set of guarded strings as indicated in
Figure~\ref{fig:interp}. A bare event $e$ is interpreted as the set of
guarded strings allowed by $\phi$ and $P$. The operation $p + q$ is
set union. The constant $0$ is the empty set.  The constant 1 is the
set of all truth assignments obeying the constraint on truth
assignments, i.e. $\{10,01\}$ in the coin example.  The operation
$\cdot$ is fusion product raised to sets: $x \cdot y$ is the set of
all defined fusion products of an element of $x$ with and element of
$y$.  The operation $*$ is Kleene star, with $x* = \cup_{i\geq 0}x^i$,
where $x^i$ is the $i$-times product of $x$ with itself and $x^0=1$.
Subsets of $1$ indicated by $t$ are also elements of $K$, and these
form the Boolean algebra of tests, denoting all atoms admitted by the
constraint $\phi$ and the boolean test $t$.  We ``overload'' $\cdot$
to further indicate boolean conjunction, $+$ to further indicate
boolean disjunction, and $0,1$ to indicate the zero and one elements
of the boolean algebra. The overbar operation denotes complement in
this Boolean algebra: $\overline t$ denotes all atoms $\alpha$ such
that $\alpha \Rightarrow \neg t$.  Other boolean operators (such as
implication $\Rightarrow$) are encoded as syntactic sugar in the
standard way. The complement operation ${p}^c$ is complement at the
level of sets of guarded strings, defined to every set the universe
$\_^*$ except those denoted by $p$. Finally, the epistemic modality
$\Diamond_a$ is interpreted using Kripke semantics, as pre-image
relative to a fixed relation $R_a$ between guarded strings,
$\Diamond_a x = \setabs{u}{\exists v.\element{v}{x} \wedge u R_a v}$.
Here $u$ and $v$ are guarded strings, while $x$ is an element of $K$.

@@

\begin{figure}
  \[\begin{array}{l >{\triangleq}c l}
  \llbracket 0 \rrbracket^{\phi,P} && \{\} \\  
  \llbracket 1 \rrbracket^{\phi,P} && \{\alpha \mid \alpha \Rightarrow \phi\}\\
  \llbracket e \rrbracket^{\phi,P}&& \{\alpha \cdot e \cdot \beta \mid
  (\alpha,\beta) \in P(e), \alpha, \beta \Rightarrow \phi \}
  \\ \llbracket \_ \rrbracket^{\phi,P}&& \bigcup_{e \in
    \dom(P)}\llbracket e \rrbracket^{\phi,P}\\
  \llbracket t \rrbracket^{\phi,P} &&  \{\alpha \mid \alpha \Rightarrow \phi \wedge t\}\\
  \llbracket{p+q}\rrbracket^{\phi,P} && \llbracket{p}\rrbracket^{\phi,P} \cup \llbracket q \rrbracket^{\phi,P}\\
  \llbracket{p \& q}\rrbracket^{\phi,P} && \llbracket{p}\rrbracket^{\phi,P} \cap \llbracket q \rrbracket^{\phi,P}\\
  \llbracket{p;q}\rrbracket^{\phi,P} && \llbracket p \rrbracket^{\phi,P} \cdot \llbracket{q}\rrbracket^{\phi,P}\\
  \llbracket p^*\rrbracket^{\phi,P} && \displaystyle\bigcup_n {\underbrace{\llbracket p \rrbracket^{\phi,P}\cdot \ldots \cdot \llbracket p \rrbracket^{\phi,P}}_{n}} \\
  \llbracket \sim p \rrbracket^{\phi,P} && \llbracket \_^*\rrbracket^{\phi,P} \setminus \llbracket p \rrbracket^{\phi,P} \\
  \llbracket \langle a \rangle~p \rrbracket && \doublehat{R_a}\llbracket p \rrbracket

@@

\section{Direct Bounded Implementation}

>>>>

Rather than guarded strings, elements of a guarded string model for
KAT are sets of guarded strings.  In the application, these elements
have the interpretation of propositions (sets of possible worlds)
and/or event types.  An event such as $u$ in the guarded string model
corresponds to the set of guarded strings where the bare event is
flanked by compatible truth assigments, as defined in an Epik program
by the event declaration and the constraint declaration.  These have
the status of event types, in that they can ``happen'' in different
possible worlds. Here happening corresponds to incrementing the world
with some compatible element of the event type.  Where $e$ is a bare
event symbol, $\llbracket e \rrbracket^{\varphi,P}$ is the corresponding
set of guarded strings, consisting of the bare event decorated with
compatible truth assignments denoted by $P$ that are admitted by the
world-constraints $\varphi$.  See this formally in
Figure~\ref{fig:interp}, and the examples in \ex{events1}.
% u (!q!p):(!qp) + (!qp):(q!p) + (q!p):(qp) 173
% d (!qp):(!q!p) + (q!p):(!qp) + (qp):(q!p)

\begin{example} \label{events1}
\(
\begin{array}[t]{l|c}
  e & \llbracket e \rrbracket^{\varphi, P} \\ \hline  
  u & \{ 00u01, 01u10, 10u11\}\\
  d & \{ 01d00, 10d01, 11d10\}\\
  a_1 & \{10 a_1 10 \}\\
  a_0 & \{01 a_0 01\}\\
  b_1 & \{10 b_1 10 \}\\
  b_0 & \{01 b_0 01 \}\\
\end{array}
\)
\end{example}

@@

We interpret each term in as a set of guarded strings as indicated in
Figure~\ref{fig:interp}. A bare event $e$ is interpreted as the set of
guarded strings allowed by $\varphi$ and $P$. The operation $p + q$ is
set union. The constant $0$ is the empty set.  The constant 1 is the
set of all truth assignments obeying the constraint on truth
assignments, i.e. $\{10,01\}$ in the coin example.  The operation
$\cdot$ is fusion product raised to sets: $x \cdot y$ is the set of
all defined fusion products of an element of $x$ with and element of
$y$.  The operation $*$ is Kleene star, with $x* = \cup_{i\geq 0}x^i$,
where $x^i$ is the $i$-times product of $x$ with itself and $x^0=1$.
Subsets of $1$ indicated by $t$ are also elements of $K$, and these
form the Boolean algebra of tests, denoting all atoms admitted by the
constraint $\varphi$ and the boolean test $t$.  We ``overload'' $\cdot$
to further indicate boolean conjunction, $+$ to further indicate
boolean disjunction, and $0,1$ to indicate the zero and one elements
of the boolean algebra. The overbar operation denotes complement in
this Boolean algebra: $\overline t$ denotes all atoms $\alpha$ such
that $\alpha \Rightarrow \neg t$.  Other boolean operators (such as
implication $\Rightarrow$) are encoded as syntactic sugar in the
standard way. The complement operation ${p}^c$ is complement at the
level of sets of guarded strings, defined to every set the universe
$\_^*$ except those denoted by $p$. Finally, the epistemic modality
$\Diamond_a$ is interpreted using Kripke semantics, as pre-image
relative to a fixed relation $R_a$ between guarded strings,
$\Diamond_a x = \setabs{u}{\exists v.\element{v}{x} \wedge u R_a v}$.
Here $u$ and $v$ are guarded strings, while $x$ is an element of $K$.

@@

\begin{figure}
  \[\begin{array}{l >{\triangleq}c l}
  \llbracket 0 \rrbracket^{\varphi,P} && \{\} \\  
  \llbracket 1 \rrbracket^{\varphi,P} && \{\alpha \mid \alpha \Rightarrow \varphi\}\\
  \llbracket e \rrbracket^{\varphi,P}&& \{\alpha \cdot e \cdot \beta \mid
  (\alpha,\beta) \in P(e), \alpha, \beta \Rightarrow \varphi \}
  \\ \llbracket \_ \rrbracket^{\varphi,P}&& \bigcup_{e \in
    \dom(P)}\llbracket e \rrbracket^{\varphi,P}\\
  \llbracket t \rrbracket^{\varphi,P} &&  \{\alpha \mid \alpha \Rightarrow \varphi \wedge t\}\\
  \llbracket{p+q}\rrbracket^{\varphi,P} && \llbracket{p}\rrbracket^{\varphi,P} \cup \llbracket q \rrbracket^{\varphi,P}\\
  \llbracket{p \& q}\rrbracket^{\varphi,P} && \llbracket{p}\rrbracket^{\varphi,P} \cap \llbracket q \rrbracket^{\varphi,P}\\
  \llbracket{p;q}\rrbracket^{\varphi,P} && \llbracket p \rrbracket^{\varphi,P} \cdot \llbracket{q}\rrbracket^{\varphi,P}\\
  \llbracket p^*\rrbracket^{\varphi,P} && \displaystyle\bigcup_n {\underbrace{\llbracket p \rrbracket^{\varphi,P}\cdot \ldots \cdot \llbracket p \rrbracket^{\varphi,P}}_{n}} \\
  \llbracket \sim p \rrbracket^{\varphi,P} && \llbracket \_^*\rrbracket^{\varphi,P} \setminus \llbracket p \rrbracket^{\varphi,P} \\
  \llbracket \langle a \rangle~p \rrbracket && \doublehat{R_a}\llbracket p \rrbracket

@@

\section{Direct Bounded Interpretation}

We also implement the semantics of Epik terms using lazy lists~\cite{LazyLists}
in Haskell, rather than the direct interpretation as sets. Unfortunately,
regular expressions, and hence also Epik, can denote infinite sets of strings,
for example the term $\_^*$. Normally regular languages are represented using a
finite coalgebra~\cite{}. However the non-distributivity of $\diamond$ accross
$;$ complicates this construction~\footnote{Axiomatic and Coalgebraic models for
  Epik are open questions}. To sidestep this, we parameterize the interpretation
function on a positive integer $n$ and only produce guarded strings of length
$n$ or less.  The length of a guarded string $g$, written $|g|$ is the number of
events in the string, so $\alpha$ has length $0$, $\alpha e\beta$ has length
$1$, and $\alpha e \beta e' \gamma$ has length $2$.

We translate the Epik terms into a Haskell algebraic datatype that represents
terms with the same signature as described in Section 2. We then paramaterize
the interpretation function, $\llparenthesis p \rrparenthesis^{\varphi,P}_n$ on an
integer $n$ that describes the maximum length of a string we will produce. The
bounded interpretation into lists of strings is very similar to the unbounded
interpretation into sets of strings, except for the bounds checking. The full
details are shown in Figure~\ref{fig:list-interp}: we use list-comprehension
notation, which is analogous to set-builder notation, except that it produces a
list (potentially with duplicated elements) instead of a set, and so is written
with square brackets $[]$. We use $+$ to indicate list concatenation, and $\sum
l$ to indicate the concatenation of a list of lists $l$. We overload the set
membership symbol $\in$ to also indicate membership in a list. We also lift the
fusion operator $\cdot$ to lists. For a list of guarded strings $l$, we write
$l|_n$ as shorthand for the set $[ g~|~g \in l , |g| \leq n]$.

First of all, whenever $n = 0$, the set of guarded strings is empty. TODO ADD
MORE TEXT.

The final step is to convert the list of guarded strings into a set of guarded
strings. Since a list is just an ordered multiset, we forget the order, and
remove all duplicates. Haskell's libraries have a built-in function for this,
\texttt{Data.Set.fromList}, which converts a list into a set.

\begin{figure}
  \[\begin{array}{r >{\triangleq}c l}
  \llparenthesis p \rrparenthesis^{\varphi,P}_0 && [] \\
  \llparenthesis 0 \rrparenthesis^{\varphi,P}_n && [] \\
  \llparenthesis 1 \rrparenthesis^{\varphi,P}_n && [\alpha \mid \alpha \leq \varphi] \\
  \llparenthesis e \rrparenthesis^{\varphi,P}_n && [\alpha \cdot e \cdot \beta \mid (\alpha, \beta) \in P(e), \alpha\vee \beta \Rightarrow \varphi] \\
  \llparenthesis \_ \rrparenthesis^{\varphi,P}_n && \sum [\llparenthesis e \rrparenthesis^{\varphi,P} \mid e \in \dom(P) ] \\
  \llparenthesis t \rrparenthesis^{\varphi,P}_n && [\alpha\mid \alpha \Rightarrow \varphi \wedge t] \\
  \llparenthesis p + q \rrparenthesis^{\varphi,P}_n && \llparenthesis p \rrparenthesis^{\varphi,P}_n + \llparenthesis q \rrparenthesis^{\varphi,P}_n \\
  \llparenthesis p \& q \rrparenthesis^{\varphi,P}_n &&
  [g~|~g \in \llparenthesis p \rrparenthesis_n^{\varphi,P}, g \in \llparenthesis q \rrparenthesis_n^{\varphi,P}] \\
  \llparenthesis p;q \rrparenthesis^{\varphi,P}_n &&
  \left(\llparenthesis p \rrparenthesis^{\varphi,P}_n \cdot \llparenthesis q\rrparenthesis^{\varphi,P}_n\right)|_n\\
  \llparenthesis p^* \rrparenthesis^{\varphi,P}_n && [] + \left(\llparenthesis p\rrparenthesis^{\varphi,P}_n \cdot  \llparenthesis p^* \rrparenthesis^{\varphi,P}_{n-i}\right)|_n \\
  \multicolumn 2 r {} & \text{where}~i = \max\{1,\min\{|g|\mid g \in \llbracket p \rrbracket^{\varphi,P}\}\}\\
  \llparenthesis \sim p \rrparenthesis^{\varphi,P}_n && [g \mid g \in \llbracket \_^* \rrbracket^{\varphi,P}_n, g \not\in \llbracket p \rrbracket^{\varphi,P}_n] \\
  \llparenthesis \langle a \rangle p \rrparenthesis^{\varphi,P}_n &&
  [g' \mid g \in g'~\doublehat{R_a}~g, g \in \llparenthesis p \rrparenthesis^{\varphi,P}]
  \end{array}
  \]
\end{figure}

TODO Describe advantages of Lazy Lists
